Concurrent Programming in Harmony

### 2021-12-25 book/paper.tex
More formatting stuff

Harmony is not an object-oriented language like Python is.  In Python,
you can pass a reference to an object to a method, and that method
can then update the object.  In Harmony, it is also sometimes convenient
to have a method update a shared variable specified as an argument.
For this, as mentioned in \autoref{ch:harmonymachine},
each shared variable has an \emph{address}, itself a Harmony value.
\index{address}%
If <{x}> is a shared variable, then the expression <{?x}> is the address of <{x}>.
If a variable contains an address, we call that variable a \emph{pointer}.
\index{pointer}%
If <{p}> is a pointer to a shared variable, then the
expression \texttt{!}<{p}> is the value of the shared variable.
In particular, <{!?x = x}>.
This is similar to how C pointers work (\texttt{*\string&}<{x = x}>).

@@

\section*{Exercises}
\begin{problems}
\item The <{put}> method you implemented in \autoref{ex:bbmesa} cannot be used
in interrupt handlers for two reasons: (1) it is not interrupt-safe, and (2)
it may block for a long time if the buffer is full.  Yet, it would be useful if,
say, a keyboard interrupt handler could place an event on a shared queue.
Implement a new method \texttt{i\_put}(<{item}>) that does not
block.  Instead, it should return <{False}> if the buffer is full and <{True}>
if the item was successfully enqueued.
The method also needs to be interrupt-safe.
\end{problems}

@@

As an example, we will revisit the producer/consumer problem.
The code in \autoref{fig:hw} is based on code developed by Herlihy and
Wing~\cite{HW87}.
The code is a ``proof of existence'' for non-blocking synchronization; it
is not necessarily practical.
There are two variables.  <{items}> is an unbounded array with each
entry initialized to <{None}>.  <{back}> is an index into the
array and points to the next slot where a new value is inserted.
The code uses two atomic operations:
\begin{itemize}
\item <{inc(p)}>: atomically increments \texttt{!}<{p}> and returns
the old value;
\item <{exch(p)}>: sets \texttt{!}<{p}> to <{None}> and returns
the old value.
\end{itemize}

@@

There are three levels of precedence.
Application has the highest precedence.  So, \texttt{!}<{a b}> is interpreted as
\texttt{!}(<{a b}>) and <{a b}> \texttt{+} <{c d}> is interpreted as
(<{a b}>) \texttt{+} (<{c d}>).
Unary operators have the next highest precedence,
and the remaining operators have the lowest precedence.
For example, $-2 + 3$ evaluates to 1, not $-5$.

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
<{tas(lk)}> & test-and-set on <{!lk}> \\
\hline
<{cas}(ptr, old, new)}> & compare-and-swap on <{!ptr}> \\
\hline
<{BinSem(v)}> & return a binary semaphore initialized to <{v}> \\
\hline
<{Lock()}> & return a binary semaphore initialized to <{False}> \\
\hline
<{acquire(bs)}> & acquire binary semaphore \texttt{!}<{bs}> \\
\hline
<{release(bs)}> & release binary semaphore \texttt{!}<{bs}> \\
\hline
<{Condition()}> & return a condition variable \\
\hline
<{wait(c, lk)}> & wait on condition variable \texttt{!}<{c}>
and lock <{lk}> \\ \hline
<{notify(c)}> & notify a thread waiting on condition variable \texttt{!}<{c}> \\
\hline
<{notifyAll(c)}> & notify all threads waiting on condition variable \texttt{!}<{c}> \\
\hline
<{Semaphore(cnt)}> & return a counting semaphore initialized to <{cnt}> \\
\hline
<{P(sema)}> & procure \texttt{!}<{sema}>  \\
\hline
<{V(sema)}> & vacate \texttt{!}<{sema}>  \\
\hline
<{Queue}>() & return a synchronized queue object \\
\hline
<{get(q)}> & return next element of <{q}>, blocking if empty \\
\hline
<{put(q, item)}> & add <{item}> to <{a}> \\
\hline
\end{tabular}

>>>>

Harmony is not an object-oriented language like Python is.  In Python,
you can pass a reference to an object to a method, and that method
can then update the object.  In Harmony, it is also sometimes convenient
to have a method update a shared variable specified as an argument.
For this, as mentioned in \autoref{ch:harmonymachine},
each shared variable has an \emph{address}, itself a Harmony value.
\index{address}%
If <{x}> is a shared variable, then the expression <{?x}> is the address of <{x}>.
If a variable contains an address, we call that variable a \emph{pointer}.
\index{pointer}%
If <{p}> is a pointer to a shared variable, then the
expression <{!p}> is the value of the shared variable.
In particular, <{!?x = x}>.
This is similar to how C pointers work (\texttt{*\string&}<{x = x}>).

@@

<{:const:i_put}>

\section*{Exercises}
\begin{problems}
\item The <{put}> method you implemented in \autoref{ex:bbmesa} cannot be used
in interrupt handlers for two reasons: (1) it is not interrupt-safe, and (2)
it may block for a long time if the buffer is full.  Yet, it would be useful if,
say, a keyboard interrupt handler could place an event on a shared queue.
Implement a new method <{i_put(item)}> that does not
block.  Instead, it should return <{False}> if the buffer is full and <{True}>
if the item was successfully enqueued.
The method also needs to be interrupt-safe.
\end{problems}

@@

As an example, we will revisit the producer/consumer problem.
The code in \autoref{fig:hw} is based on code developed by Herlihy and
Wing~\cite{HW87}.
The code is a ``proof of existence'' for non-blocking synchronization; it
is not necessarily practical.
There are two variables.  <{items}> is an unbounded array with each
entry initialized to <{None}>.  <{back}> is an index into the
array and points to the next slot where a new value is inserted.
The code uses two atomic operations:
\begin{itemize}
\item <{inc(p)}>: atomically increments <{!p}> and returns
the old value;
\item <{exch(p)}>: sets <{!p}> to <{None}> and returns the old value.
\end{itemize}

@@

There are three levels of precedence.
Application has the highest precedence.  So, <{!a b}> is interpreted as
<{!(a b)}> and <{a b + c d}> is interpreted as <{(a b) + (c d)}>.
Unary operators have the next highest precedence,
and the remaining operators have the lowest precedence.
For example, $-2 + 3$ evaluates to 1, not $-5$.

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
<{tas(lk)}> & test-and-set on <{!lk}> \\
\hline
<{cas}(ptr, old, new)}> & compare-and-swap on <{!ptr}> \\
\hline
<{BinSem(v)}> & return a binary semaphore initialized to <{v}> \\
\hline
<{Lock()}> & return a binary semaphore initialized to <{False}> \\
\hline
<{acquire(bs)}> & acquire binary semaphore <{!bs}> \\
\hline
<{release(bs)}> & release binary semaphore <{!bs}> \\
\hline
<{Condition()}> & return a condition variable \\
\hline
<{wait(c, lk)}> & wait on condition variable <{!c}>
and lock <{lk}> \\ \hline
<{notify(c)}> & notify a thread waiting on condition variable <{!c}> \\
\hline
<{notifyAll(c)}> & notify all threads waiting on condition variable <{!c}> \\
\hline
<{Semaphore(cnt)}> & return a counting semaphore initialized to <{cnt}> \\
\hline
<{P(sema)}> & procure <{!sema}>  \\
\hline
<{V(sema)}> & vacate <{!sema}>  \\
\hline
<{Queue}>() & return a synchronized queue object \\
\hline
<{get(q)}> & return next element of <{q}>, blocking if empty \\
\hline
<{put(q, item)}> & add <{item}> to <{a}> \\
\hline
\end{tabular}
